---
title: "CDC Death Counts"
output:
  html_document:
    df_print: paged
---

# Load Weekly CDC Death Counts

```{r setup}


library(tidyverse)
library(lubridate)
library(readODS)
library(qcc)

week_deaths <- read.csv("Weekly_Counts_of_Deaths_by_State_and_Select_Causes__2019-2020.csv")
glimpse(week_deaths)

```

# Weekly Natural Causes

Have deaths from all natural causes increased?

## Running Record

```{r}

US.summ <- week_deaths %>%
  mutate(Week.Ending.Date = ymd(Week.Ending.Date)) %>% 
  group_by(Week.Ending.Date) %>%
  summarise(US_Natural = sum(Natural.Cause),
            US_COVID_multi = sum(COVID.19..U071..Multiple.Cause.of.Death.),
            US_COVID_under = sum(COVID.19..U071..Underlying.Cause.of.Death.)
            )

qplot(Week.Ending.Date, US_Natural, data = US.summ, geom = "path")

```

Large spike in deaths by Natural causes in April 2020, another in August-September. What if we remove New York and New Jersey?

```{r}

US.less2.summ <- week_deaths %>%
  filter(!(Jurisdiction.of.Occurrence %in% c("New York", "New Jersey"))) %>% 
  mutate(Week.Ending.Date = ymd(Week.Ending.Date)) %>% 
  group_by(Week.Ending.Date) %>%
  summarise(US_Natural = sum(Natural.Cause),
            US_COVID_multi = sum(COVID.19..U071..Multiple.Cause.of.Death.),
            US_COVID_under = sum(COVID.19..U071..Underlying.Cause.of.Death.)
            )

qplot(Week.Ending.Date, US_Natural, data = US.less2.summ, geom = "path")

```

Excluding the two States with by far the largest COVID deaths per capita did not change the shape of the plot very much.

### Shewhart Chart

```{r}

ok.di.df <- covid.us.df %>%
  filter(state == "OK") %>%
  arrange(date)

ok.di.df %>% 
  select(date, deathIncrease, death)

```

March 28, 2020 is the first date with more than 2 deaths, and also the date on which cumulative deaths exceeded 10. We will start the Shewhart chart there.

```{r}

ok.di.df <- ok.di.df %>% 
  filter(death >= 10)

qcc(ok.di.df$deathIncrease, type = "xbar.one", labels = as.character(ok.di.df$date))

```

Except for a spike on `r ok.di.df$date[ok.di.df$deathIncrease > 17]`, daily COVID-19 deaths in Oklahoma have varied randomly between 0 and 17, with a mean of about 6.

## Daily Totals of Hospitalized Patients

These are the counts each day of COVID-19 patients hospitalized on that day.

### Running Record

```{r}

covid.us.df %>%
  filter(state == "OK") %>%
  arrange(date) %>%
  qplot(date, hospitalizedCurrently  , data = ., geom = "path")

```

## Daily Totals of Patients in Intensive Care

These are the counts each day of COVID-19 patients in intensive care units.

### Running Record

```{r}

covid.us.df %>%
  filter(state == "OK") %>%
  arrange(date) %>%
  qplot(date, inIcuCurrently  , data = ., geom = "path")

```

# Minnesota Deaths

```{r}
covid.us.df %>%
  filter(state == "MN") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```


```{r}

mn.di.df <- covid.us.df %>%
  filter(state == "MN", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

  qcc(mn.di.df$deathIncrease, type = "xbar.one", labels = as.character(mn.di.df$date))

```

# Mississippi Deaths

```{r}
covid.us.df %>%
  filter(state == "MS") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```

```{r}

ms.di.df <- covid.us.df %>%
  filter(state == "MS", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

qcc(ms.di.df$deathIncrease, type = "xbar.one", labels = as.character(ms.di.df$date))

```

# Oregon Deaths

```{r}
covid.us.df %>%
  filter(state == "OR") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```

```{r}

or.di.df <- covid.us.df %>%
  filter(state == "OR", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

qcc(or.di.df$deathIncrease, type = "xbar.one", labels = as.character(or.di.df$date))

```

# Florida Deaths

```{r}
covid.us.df %>%
  filter(state == "FL") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```

```{r}

fl.di.df <- covid.us.df %>%
  filter(state == "FL", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

qcc(fl.di.df$deathIncrease, type = "xbar.one", labels = as.character(fl.di.df$date))

```

# All States

```{r}

covid.us.df %>%
  arrange(date) %>%
  group_by(date) %>% 
  summarize(deaths = sum(deathIncrease, na.rm = TRUE), 
            cum.deaths = sum(death, na.rm = TRUE)) %>% 
  qplot(date, deaths , data = ., geom = "path")


```

```{r}

us.di.df <- covid.us.df %>%
  group_by(date) %>%
  summarize(deaths = sum(deathIncrease, na.rm = TRUE),
            cum.deaths = sum(death, na.rm = TRUE)) %>% 
  filter(cum.deaths >= 1000) %>%
  select(date, deaths)

qcc(us.di.df$deaths, type = "xbar.one", labels = as.character(us.di.df$date))

short.us <- us.di.df %>% 
  filter(date >= ymd("2020-04-07"))
  
qcc(short.us$deaths, type = "xbar.one", labels = as.character(short.us$date))

```

