---
title: "COVID-19 Tracking"
output:
  html_document:
    df_print: paged
---

# Retrieve US COVID-19 Tracking Figures

The site https://covidtracking.com provides and API. I will download the "States Historical Data" file.

```{r}

library(jsonlite)
library(tidyverse)
library(lubridate)
library(qcc)

# covid.us.df <- fromJSON("response_1588546050977.json") %>%
#   mutate(date = ymd(date))

covid.us.df <- fromJSON("https://covidtracking.com/api/v1/states/daily.json") %>%
  mutate(date = ymd(date))

glimpse(covid.us.df)

```

# Oklahoma COVID-19 History

## Daily Death Counts

### Running Record

```{r}

covid.us.df %>%
  filter(state == "OK") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```

### Shewhart Chart

```{r}

ok.di.df <- covid.us.df %>%
  filter(state == "OK") %>%
  arrange(date)

ok.di.df %>% 
  select(date, deathIncrease, death)

```

March 28, 2020 is the first date with more than 2 deaths, and also the date on which cumulative deaths exceeded 10. We will start the Shewhart chart there.

```{r}

ok.di.df <- ok.di.df %>% 
  filter(death >= 10)

qcc(ok.di.df$deathIncrease, type = "xbar.one", labels = as.character(ok.di.df$date))

```

Except for a spike on `r ok.di.df$date[ok.di.df$deathIncrease > 17]`, daily COVID-19 deaths in Oklahoma have varied randomly between 0 and 17, with a mean of about 6.

## Daily Totals of Hospitalized Patients

These are the counts each day of COVID-19 patients hospitalized on that day.

### Running Record

```{r}

covid.us.df %>%
  filter(state == "OK") %>%
  arrange(date) %>%
  qplot(date, hospitalizedCurrently  , data = ., geom = "path")

```

## Daily Totals of Patients in Intensive Care

These are the counts each day of COVID-19 patients in intensive care units.

### Running Record

```{r}

covid.us.df %>%
  filter(state == "OK") %>%
  arrange(date) %>%
  qplot(date, inIcuCurrently  , data = ., geom = "path")

```

# Minnesota Deaths

```{r}
covid.us.df %>%
  filter(state == "MN") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```


```{r}

mn.di.df <- covid.us.df %>%
  filter(state == "MN", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

  qcc(mn.di.df$deathIncrease, type = "xbar.one", labels = as.character(mn.di.df$date))

```

# Mississippi Deaths

```{r}
covid.us.df %>%
  filter(state == "MS") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```

```{r}

ms.di.df <- covid.us.df %>%
  filter(state == "MS", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

qcc(ms.di.df$deathIncrease, type = "xbar.one", labels = as.character(ms.di.df$date))

```

# Oregon Deaths

```{r}
covid.us.df %>%
  filter(state == "OR") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```

```{r}

or.di.df <- covid.us.df %>%
  filter(state == "OR", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

qcc(or.di.df$deathIncrease, type = "xbar.one", labels = as.character(or.di.df$date))

```

# Florida Deaths

```{r}
covid.us.df %>%
  filter(state == "FL") %>%
  arrange(date) %>%
  qplot(date, deathIncrease  , data = ., geom = "path")

```

```{r}

fl.di.df <- covid.us.df %>%
  filter(state == "FL", death >= 10) %>%
  arrange(date) %>%
  select(date, deathIncrease)

qcc(fl.di.df$deathIncrease, type = "xbar.one", labels = as.character(fl.di.df$date))

```

# All States

```{r}

covid.us.df %>%
  arrange(date) %>%
  group_by(date) %>% 
  summarize(deaths = sum(deathIncrease, na.rm = TRUE), 
            cum.deaths = sum(death, na.rm = TRUE)) %>% 
  qplot(date, deaths , data = ., geom = "path")


```

```{r}

us.di.df <- covid.us.df %>%
  group_by(date) %>%
  summarize(deaths = sum(deathIncrease, na.rm = TRUE),
            cum.deaths = sum(death, na.rm = TRUE)) %>% 
  filter(cum.deaths >= 1000) %>%
  select(date, deaths)

qcc(us.di.df$deaths, type = "xbar.one", labels = as.character(us.di.df$date))

short.us <- us.di.df %>% 
  filter(date >= ymd("2020-04-07"))
  
qcc(short.us$deaths, type = "xbar.one", labels = as.character(short.us$date))

```

